A Transactional Memory System for Parallel Python

### 2014-07-30 talk/dls2014/paper/paper.tex
clarifications

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best for the embarrassingly
parallel ones ($avg=2.6\times$ speedup) and a little less for
the others ($avg=2.0\times$ speedup). The reason for this difference is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

Looking at the average overhead from switching from GIL to STM, we see
that it is $\approx 43.3\%$. The maximum in richards is $71\%$. In all
benchmarks \emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on
two threads despite of this overhead.  The achieved speedup comparing
STM to the GIL is between $1.14\times$ and $1.94\times$.

@@

Since PyPy comes with a JIT~\cite{cfbolz09} to make its overhead
compared to CPython go away, we will now look at how well STM works
together with it.

@@

The speedups from enabling the JIT in these benchmarks range from
$10-50\times$. This is why we had to do without CPython here, since it
would be much further up in the plots. Also, to make jitting
code worthwhile, we increased the input size of all benchmarks to get
reasonable execution times (denoted with ``(large)'' in benchmark names).

@@

The results are presented in Figure~\ref{fig:performance-jit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around $1-2.4\times$, and we beat GIL performance
in half of the benchmarks.

>>>>

On 4 cores, PyPy using our STM system (\emph{pypy-stm-nojit}) scales
in all benchmarks to a certain degree. It scales best for the
embarrassingly parallel ones ($avg=2.6\times$ speedup over 1 thread)
and a little less for the others ($avg=2.0\times$ speedup over 1
thread). The reason for this difference is
that in the former group there are no real, logical conflicts -- all
threads do independent calculations. STM simply replaces the GIL in
those programs. In the latter group, the threads work on a common data
structure and therefore create much more conflicts, which limits the
scalability. Here we make active use of the STM-supported atomic
blocks to synchronise access to the data structure. The hope
is that the STM system is still able to parallelise, even if we use
the atomic blocks in a coarse-grained way. While less than the other
benchmarks, we indeed see some speedup going from 1 to 3 threads.
There is no visible benefit from 3 to 4 threads, even a slight
regression.

Looking at the average overhead on a single thread that is induced by
switching from GIL to STM, we see that it is $\approx 43.3\%$. The
maximum in richards is $71\%$. In all benchmarks \emph{pypy-stm-nojit}
beats \emph{pypy-nojit} already on two threads, despite of this
overhead.  The achieved speedup comparing the lowest runtimes of STM
to the single-threaded GIL execution is between $1.14\times$ and
$1.94\times$.

@@

Since PyPy comes with a JIT~\cite{cfbolz09} to make it vastly faster
than CPython, we will now look at how well STM works together with it.

@@

The speedups from simply enabling the JIT in these benchmarks range from
$10-50\times$. This is why we had to do without CPython here, since it
would be much further up in the plots. Also, to make jitting
code worthwhile, we increased the input size of all benchmarks to get
reasonable execution times (denoted with ``(large)'' in benchmark names).

@@

The results are presented in Figure~\ref{fig:performance-jit}. We see
that the performance gains are much less reliable.  The slowdown
factor for switching from GIL to STM ranges around $1-2.4\times$, and
we beat the GIL's single-thread performance in half of the benchmarks.
